Application of a reaction-diffusion equation to the spread of the invasive ascidian *Botrylloides violaceus*

### 2024-04-29 PDEs/math245_hw6.qmd
presentation

@@

The mean value property says that for a harmonic function $u$ in a disk $D$, continuous in its closure $\bar{D}$, the value of $u$ at the center of $D$ equals the average of $u$ on its circumference. The average value of a continuous function $f(x)$ over the interval $[a,b]$ is given by 
$$f_{avg}(x)=\frac{1}{b-a}\int_a^b f(x)\;dx$$
Here,
\begin{align*}
u_{avg}(\theta)& =\frac{1}{2\pi}\int_0^{2\pi} (\sin\theta+1)\;d\theta \\
&=\frac{1}{2\pi}\Bigl[(\cos\theta+\theta) \Bigr]_0^{2\pi} \\
&=\frac{1}{2\pi}\Bigl[(\cos{2\pi}+2\pi)-\cos{0} \Bigr] \\
&= \frac{1}{2\pi}(1+2\pi-1)=1
\end{align*}
Therefore, the value of $u$ at the origin is 1.

@@

Since (\ref{eqn:Green}) must hold for any $(u,v)$, take $v=1$. Then $\Delta v = 0 = \frac{\partial v}{\partial n}$, so
\begin{align*}
\iint\limits_{D}(u\Delta v-v\Delta u)\,dA &=\int\limits_{\partial D} \left(u\frac{\partial v}{\partial n} - v\frac{\partial u}{\partial n}\right)\,ds \quad \text{becomes} \\
\iint\limits_{D}(0-\Delta u)\,dA &=\int\limits_{\partial D} \left(0 - \frac{\partial u}{\partial n}\right)\,ds \\
0 &=-\int\limits_{\partial D} g\left( \theta \right)\,ds
\end{align*}

Where we used the fact that by (\ref{eqn:prob5}), $\Delta u = 0$ and $\frac{\partial u}{\partial n}=u_r=g(\theta)$. Thus, the necessary condition for the existence of a solution is $$\int\limits_{\partial D} g\left( \theta \right)\,ds=0$$

@@

$$u(r,\theta)=A_0+\sum_{n=1}^{\infty}r^{-n} \left(A_n\cos{(n\theta)}+B_n\sin{(n\theta)}\right) $$
$$u_r(r,\theta)=A_0+\sum_{n=1}^{\infty}-nr^{-n-1} \left(A_n\cos{(n\theta)}+B_n\sin{(n\theta)}\right) $$

Finally, we use the inhomogeneous BC at $r=a$. Setting $r=a$ in the series above, we require that
$$g(\theta)=A_0+\sum_{n=1}^{\infty}-na^{-n-1} \left(A_n\cos{(n\theta)}+B_n\sin{(n\theta)}\right) $$

@@

$$A_0=\frac{1}{2\pi} \int_{-\pi}^{\pi}g(\theta)\,d\theta $$
But from the necessary condition for the existence of a solution, this integral must be zero.
Therefore, our solution to (\ref{eqn:prob5}) is
\begin{tcolorbox}[colback=blue!5!white,colframe=blue!75!black,title=Solution]

$$u(r,\theta)=\sum_{n=1}^{\infty}r^{-n} \left(A_n\cos{(n\theta)}+B_n\sin{(n\theta)}\right) $$
where
$$A_n=\frac{-1}{n\pi(a^{n+1})} \int_{-\pi}^{\pi}g(\theta) \cos{(n\theta)}\,d\theta $$
$$B_n=\frac{-1}{n\pi(a^{n+1})} \int_{-\pi}^{\pi}g(\theta) \sin{(n\theta)}\,d\theta $$
\end{tcolorbox}

@@

Now set $u=v=w$:

@@

But from (\ref{eqn:unique2}), $\Delta w=\frac{\partial w}{\partial n}=0$, so

>>>>

# (Wednesday morning correction to Problem 5)

@@

The mean value property says that for a harmonic function $u$ in a disk $D$, continuous in its closure $\bar{D}$, the value of $u$ at the center of $D$ equals the average of $u$ on its circumference. The average value of a continuous function $f(x)$ over the interval $[a,b]$ is given by 
$$f_{avg}(x)=\frac{1}{b-a}\int_a^b f(x)\;dx$$
Here,
\begin{align*}
u(0) &=\text{average of u over } \partial{B}(0,2) \\
&=\frac{1}{2\pi R} \int_{\partial{B}(0,R)} u \; ds \\
&=\frac{1}{4\pi} \int_{\partial{B}(0,R)} u \; ds \\
&=\frac{1}{4\pi} \int_{\theta=0}^{\theta=2\pi} u \cdot R \; d\theta \\
& =\frac{1}{2\pi}\int_0^{2\pi} u\;d\theta \\
& =\frac{1}{2\pi}\int_0^{2\pi} (\sin\theta+1)\;d\theta \\
&=\frac{1}{2\pi}\Bigl[(\cos\theta+\theta) \Bigr]_0^{2\pi} \\
&=\frac{1}{2\pi}\Bigl[(\cos{2\pi}+2\pi)-\cos{0} \Bigr] \\
&= \frac{1}{2\pi}(1+2\pi-1)=1
\end{align*}
Therefore, the value of $u$ at the origin is 1.

@@

Since (\ref{eqn:Green}) must hold for any $(u,v)$, take $v=1$. Then $\Delta v = 0 = \frac{\partial v}{\partial n}$, so
\begin{align*}
\iint\limits_{D}(u\Delta v-v\Delta u)\,dA &=\int\limits_{\partial D} \left(u\frac{\partial v}{\partial n} - v\frac{\partial u}{\partial n}\right)\,ds \quad \text{becomes} \\
\iint\limits_{D}(0-\Delta u)\,dA &=\int\limits_{\partial D} \left(0 - \frac{\partial u}{\partial n}\right)\,ds \\
0 &=-\int\limits_{\partial D} g\left( \theta \right)\,d\theta
\end{align*}

Where we used the fact that by (\ref{eqn:prob5}), $\Delta u = 0$ and $\frac{\partial u}{\partial n}=u_r=g(\theta)$. Thus, the necessary condition for the existence of a solution is $$\int\limits_{\partial D} g\left( \theta \right)\,d\theta=0$$

@@

$$u(r,\theta)=A_0+\sum_{n=1}^{\infty}r^{-n} \left(A_n\cos{(n\theta)}+B_n\sin{(n\theta)}\right) $$
$$u_r(r,\theta)=\sum_{n=1}^{\infty}-nr^{-n-1} \left(A_n\cos{(n\theta)}+B_n\sin{(n\theta)}\right) $$

Finally, we use the inhomogeneous BC at $r=a$. Setting $r=a$ in the series above, we require that
$$g(\theta)=\sum_{n=1}^{\infty}-na^{-n-1} \left(A_n\cos{(n\theta)}+B_n\sin{(n\theta)}\right) $$

@@

$$-nA_na^{-n-1}=\frac{1}{\pi} \int_{-\pi}^{\pi}g(\theta)\cos(n\theta)\,d\theta $$
$$-nB_na^{-n-1}=\frac{1}{\pi} \int_{-\pi}^{\pi}g(\theta)\sin(n\theta)\,d\theta $$

Therefore, our solution to (\ref{eqn:prob5}) is
\begin{tcolorbox}[colback=blue!5!white,colframe=blue!75!black,title=Solution]

$$u(r,\theta)=A_0+\sum_{n=1}^{\infty}r^{-n} \left(A_n\cos{(n\theta)}+B_n\sin{(n\theta)}\right) $$
where
$$A_n=-\frac{a^{n+1}}{n\pi} \int_{-\pi}^{\pi}g(\theta) \cos{(n\theta)}\,d\theta $$
$$B_n=-\frac{a^{n+1}}{n\pi} \int_{-\pi}^{\pi}g(\theta) \sin{(n\theta)}\,d\theta $$
\end{tcolorbox}

@@

Now replace $u$ and $v$ in this formula with $w$:

@@

But from (\ref{eqn:unique2}), $\Delta w=0$ in $D$ and $\frac{\partial w}{\partial n}=0$ on $\partial{D}$, so

### 2024-04-29 PDEs/paper_outline.qmd
presentation

---
title: "Application of a reaction-diffusion equation to the spread of the invasive ascidian *Botrylloides violaceus*"
author: "Ruby Krasnow"
date: "2024-04-22"
format:
  pdf:
    documentclass: scrartcl
    include-in-header:
      - text: \usepackage{booktabs, caption, longtable, colortbl, array, graphicx}
execute:
  echo: false
bibliography: references.bib
csl: molecular-ecology.csl
link-external-newwindow: true 
---

@@

@@

\begin{equation}\label{eqn:Fisher}
\frac{\partial{u}}{\partial{t}}=
\text{D}\frac{\partial{u}^2}{\partial{x^2}}+r(1-\frac{u}{K})
\end{equation}

@@

Now, we can quantify the spatial spread of the population by considering how quickly a level set $U(x,t)=U_c$ moves in space, where $U_c$ could be considered a threshold of detection. In other words, below $U_c$ it is so difficult to find the invasive organisms that for practical purposes they are not present. By rearranging (\ref{eqn:PDEsol}), we can get the following expression $$\left(\frac{x}{t}\right) ^2=4rD+\frac{4r\sigma^2}{t}-\frac{4(\sigma^2+Dt)}{t^2}\ln{  \left(\frac{U_{cr}\sqrt{4\pi(\sigma^2+Dt)}}{G}\right) }$$

>>>>

---
title: "Application of a reaction-diffusion equation to the spread of the invasive ascidian *Botrylloides violaceus*"
author: "Ruby Krasnow"
date: "2024-04-29"
format:
  pdf:
    documentclass: scrartcl
    include-in-header:
      - text: \usepackage{booktabs, caption, longtable, colortbl, array, graphicx}
execute:
  echo: false
bibliography: references.bib
csl: molecular-ecology.csl
link-external-newwindow: true 
---

@@

# Abstract
Modeling is a critical tool for minimizing the harmful effects of invasive species on natural ecosystems. Despite the recent popularity of correlative environmental niche models, mechanistic reaction-diffusion models remain powerful tools for predicting and controlling the spread of invasive species. The colonial ascidian *Botrylloides violaceus* has rapidly spread across the Gulf of Maine, outcompeting and smothering native species and presenting a major nuisance for shellfish aquaculture.Here, we apply arelationship between the rate of spread of an invasive species and select biological parameters(namely, the population growth rate and diffusion coefficient) derived from the generalized reaction-diffusion equation. We validated the predicted rate of spread for *B. violaceus* against rates calculated from species occurrence data from the citizen science platform iNaturalist. The model predicted a mean speed of ~49 km/year, which was extremely close to the observed mean of ~40 km/year. Although simple reaction-diffusion models have limited biological realism, the accuracy of our results suggest that estimating spreading speed based on empirical data on local
diffusion potentials and population growth rates could be an underutilized approach in invasion ecology.

@@

\begin{equation}\label{eqn:Fisher}
\frac{\partial{u}}{\partial{t}}=
\text{D}\frac{\partial{u}^2}{\partial{x^2}}+ru\left(1-\frac{u}{K}\right)
\end{equation}

@@

Now, we can quantify the spatial spread of the population by considering how quickly a level set $U(x,t)=U_c$ moves in space, where $U_c$ could be considered a threshold of detection. In other words, below $U_c$ it is so difficult to find the invasive organisms that for practical purposes they are not present. By rearranging (\ref{eqn:PDEsol}), we can get the following expression $$\left(\frac{x}{t}\right) ^2=4rD+\frac{4r\sigma^2}{t}-\frac{4(\sigma^2+Dt)}{t^2}\ln{  \left(\frac{U_{c}\sqrt{4\pi(\sigma^2+Dt)}}{G}\right) }$$
